Concrete Semantics for Pushdown Analysis:\\The Essence of Summarization

### 2013-05-10 paper.tex
Incorporated some of Stephen's comments.

\abstract{Lorem Ipsum}

@@

Programs in higher-order languages heavily use function calls and
method dispatch as part of their control flow. Until recently, flow
analyses for higher-order languages could not handle return flow precisely
\citep{ianjohnson:vardoulakis-lmcs11, ianjohnson:earl2010pdcfa}, which
leads to several spurious paths (and thus false positives) due to
their pervasiveness. These works, called CFA2 and PDCFA respectively,
use pushdown automata as their approximation's target model of
computation. They are hence called ``pushdown analyses.''\footnote{We
  will refer to the classic finite model analyses as ``regular
  analyses'' after regular languages.} The precision and performance
gains of these analyses are so beneficial that we believe new analyses
going forward should be using pushdown techniques. CFA2 and PDCFA have
difficult details to easily apply to an off-the-shelf semantics ---
especially if they feature non-local control transfer that breaks the
pushdown model. \\

This paper gives a common, simple framework to discuss both CFA2 and
PDCFA along with a recipe to apply pushdown analysis techniques to an
off-the-shelf semantics in a manner similar to
~\citet{dvanhorn:VanHorn2010Abstracting} (a technique called
abstracting abstract machines, or AAM). We then extend the state of
the art by giving a novel analysis for delimited, composable control. \\

@@

@@

A downside of the work providing \rackett{call/cc} is that it was an
algorithmic change to the already complicated CFA2 --- there was no
recipe for how to do this for one's favorite control operator. This
paper seeks to do just that with an operational view of what
summarization is, in essence. In other words, we give a concrete
semantics to the tricks that the analyses in the literature use in the
abstract, and maintain the original meaning of the language. We also
give intuitive analogies to well-established ideas/techniques so that
you too can write a pushdown analysis for your language. In order to
demonstrate the applicability of this viewpoint, we show a new
analysis for a language with composable control. All of the semantics
modeled in this paper are implemented in full detail in PLT Redex and
available at

\begin{center}
  \url{http://github.com/ianj/concrete-summaries}
\end{center}

There are common underpinnings of PDCFA and CFA2 that can be embodied
as concrete machinery in the programming language semantics: breaking
the recursive structure of continuations by indirecting them through a
table with appropriately precise keys, and memoization.  Once the
machine semantics is in this form, simple pointwise abstraction leads
to the summarization algorithms that we see in the literature. In
effect, we give a \emph{concrete} semantics to pushdown analysis.

@@

The $\Addr$ space is what controls the precision of the model. For a
concrete semantics, we require the allocation metafunction $\alloc :
\State \to \Addr$ to return fresh addresses ($\maddr \notin
\dom(\mstore)$), but any choice of address is sound.

@@

\paragraph{Tracking return points:} the magic of the method is in
keeping a table of (local) continuations for each function and store
pair. This closely mirrors the technique of store-allocating
continuations used in AAM. Instead of deferring to the table (store)
for the remainder of a continuation for each frame, we only have
indirections at function call boundaries. Note that since AAM is about
finitizing the state space, this step itself fits within the AAM
method, since continuations within functions truncated at call
boundaries are bounded by the nesting depth of those functions,
thereby making the continuation space finite. In the monovariant case,
the number of continuations is still linear in the size of the
program.

@@

\paragraph{The role of summaries:} notice the additional $\Memo$
component. A ``summary edge'' in CFA2 or equivalently an
$\epsilon$-edge in PDCFA is an edge from the source of a push edge to
the target of the matching pop edge. ``Matching'' here means there is
a path through machine reductions that don't change the stack, or
through summary edges. In our model, we only ``push'' when we call a
function, and we only ``pop'' when we return with a value. There is an
analogy to something more operational: summary edges embody
memoization. Instead of following an entire path through a call to
return a value, we simply jump from the call to the return (the result
of the call). We will return to this analogy and what it
means in the presence of first-class control in section \ref{sec:sr}.

\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
% variable lookup
    $\tpl{(\svar\mvar, \menv), \mstore, \mkont, \mktab, \mmemo}$
    &
    $\tpl{\mval,\mstore,\mkont, \mktab, \mmemo}$ if $\mval \in \mstore(\menv(\mvar))$
    \\
% application
    $\tpl{(\sapp{\mexpri0}{\mexpri1}, \menv), \mstore, \mkont, \mktab, \mmemo}$
    &
    $\tpl{(\mexpri0, \menv), \mstore, \kar{\mexpri1,\menv,\mkont}, \mktab, \mmemo}$
    \\
% argument evaluation
    $\tpl{\mval, \mstore, \kar{\mexpr,\menv,\mkont}, \mktab, \mmemo}$
    &
    $\tpl{(\mexpr, \menv), \mstore, \kfn{\mval, \mkont}, \mktab, \mmemo}$
    \\
% function call
    $\tpl{\mval,\mstore,\kfn{\vclo{\slam{\mvar}{\mexpr}}{\menv}, \mkont},\mktab,\mmemo}$
    & % actually do call,
    $\tpl{\mpoint,
          \mstore',
          \krt{\mpoint, \mstore'},
          \joinone{\mktab}{(\mpoint, \mstore')}{\mkont},
          \mmemo}$
\\
    & $\tpl{\mval, \mstore', \mkont, \mktab, \mmemo} \text{ if } \mval \in \mmemo(\mpoint,\mstore')$
    \\ % or, lookup in table
%    & $\tpl{\mval, \mstore', \mkont,\mktab,\mmemo}$ if $\mval \in \mmemo(\mpoint, \mstore')$ \\    
    where & $\mpoint = (\mexpr, \extm{\menv}{\mvar}{\maddr})$ \\
          & $\mstore' = \joinone{\mstore}{\maddr}{\mval}$
    \\
% return
    $\tpl{\mval, \mstore, \krt{\mpoint,\mstore'}, \mktab, \mmemo}$
    &
    $\tpl{\mval, \mstore, \mkont, \mktab, \joinone{\mmemo}{(\mpoint, \mstore')}{\mval}}$
    if $\mkont \in \mktab(\mpoint, \mstore')$
  \end{tabular}
  \caption{The summarizing tabular stack machine}
  \label{fig:summary-semantics}
\end{figure}

@@

The immediately related work is that of PDCFA
\citep{ianjohnson:earl2010pdcfa},
CFA2~\citep{ianjohnson:vardoulakis-lmcs11,
  ianjohnson:Vardoulakis2011Pushdown}, and AAM
~\citep{dvanhorn:VanHorn2010Abstracting}, the first two of which we
recreated in full detail. The version of CFA2 that handles
\rackett{call/cc} does not handle composable control, and is dependent
on a restricted CPS representation. They also had no way to tune the
precision of their first class continuation approximation. Since
\rackett{call/cc} can be simulated with shift/reset, this work
supercedes theirs.

@@

@@

Our goal in the future is to show that this technique is even more
widely applicable than shift and reset, but indeed to the most
intricate control operators in production languages such as in
\citet{ianjohnson:Flatt:2007:ADC:1291151.1291178}. The control
structures there are difficult to model even with AAM's techniques,
but once we can tackle AAM, there should be a straightforward route to
extending it with the pushdown techniques of this paper.

There is also the question of clients of these analyses. Most
obviously we would want to know where we can implement first-class
control more efficiently. On top of optimizations, there are security
analyses. Greater control of the stack gives us the ability to drill
deep into context-sensitive security properties and make precise
predictions.

>>>>

\begin{abstract}
  Lorem Ipsum
\end{abstract}

@@

Programs in higher-order languages heavily use function calls and
method dispatch as part of their control flow. Until recently, flow
analyses for higher-order languages could not handle return flow
precisely \citep{ianjohnson:vardoulakis-lmcs11,
  ianjohnson:earl2010pdcfa}, which leads to several spurious paths
(and thus false positives) due to the pervasiveness of function/method
calls and subsequent returns. These works, called CFA2 and PDCFA
respectively, use pushdown automata as their approximation's target
model of computation. They are hence called ``pushdown
analyses.''\footnote{We will refer to the classic finite model
  analyses as ``regular analyses'' after regular languages.} CFA2 and
PDCFA have difficult details to easily apply to an off-the-shelf
semantics --- especially if they feature non-local control transfer
that breaks the pushdown model.

There is a systematic process for transforming off-the-shelf
programming language semantics into a form amenable to \emph{regular}
analysis that has been widely applied with great success to production
programming languages ~\citet{dvanhorn:VanHorn2010Abstracting} (a
technique called abstracting abstract machines, or AAM). A
contribution of this paper is a systematic process to construct
\emph{pushdown} analyses of programming languages, due to the
precision (and often performance) benefits.

@@

This paper gives a common, simple framework to derive both CFA2 and
PDCFA using a recipe to apply pushdown analysis techniques to
arbitrary semantics in a manner similar to AAM. To exercise the recipe further, 
we give a novel analysis for delimited, composable control. \\

@@

A downside of the work providing \rackett{call/cc} is that it was an
algorithmic change to the already complicated CFA2 --- there was no
recipe for how to do this for one's favorite control operator. This
paper seeks to do just that with an operational view of what
summarization is, in essence. In other words, we give a concrete
semantics to the tricks that the analyses in the literature use in the
abstract, and maintain the original meaning of the language. We also
give intuitive analogies to well-established ideas/techniques so that
the working semanticist can write a pushdown analysis for her
language. In order to demonstrate the applicability of this viewpoint,
we show a new analysis for a language with composable control. All of
the semantics modeled in this paper are implemented in full detail in
PLT Redex~\citep{ianjohnson:Felleisen:2009:SEP:1795772} and available
online\footnote{\url{http://github.com/ianj/concrete-summaries}}.

There are common underpinnings of PDCFA and CFA2 that can be embodied
as concrete machinery in the programming language semantics: 
\begin{enumerate}
\item{breaking the recursive structure of continuations by indirecting them through a
table with appropriately precise keys, and}
\item{memoization.}
\end{enumerate}
Once the machine semantics is in this form, simple pointwise
abstraction leads to the summarization algorithms that we see in the
literature.

@@

The $\Addr$ space is what controls the precision of the model. For a
concrete semantics, we require the allocation metafunction $\alloc :
\State \to \Addr$ to return fresh addresses ($\maddr \notin
\dom(\mstore)$), but any choice of address is sound. This metafunction
along with the commitment to having no recursive datastructures is the
key to the AAM technique. All recursion can be expressed by
indirecting through addresses into a store --- tying Landin's knot. We
borrow this technique on top of our own.

@@

\paragraph{Tracking return points:} the magic of the method is in
keeping a table of (local) continuations for each function and store
pair. This closely mirrors the technique of store-allocating
continuations used in AAM. Instead of deferring to the table (store)
for the remainder of a continuation for each frame, we only have
indirections at function call boundaries (see rules 4 and 5 in figure
\ref{fig:summary-semantics} for saving and restoring
continuations). Note that since AAM is about finitizing the state
space, this step itself fits within the AAM method, since
continuations within functions truncated at call boundaries are
bounded by the nesting depth of those functions, thereby making the
continuation space finite. In the monovariant case, the number of
continuations is still linear in the size of the program.

@@

\paragraph{The role of summaries:} notice the additional $\Memo$
component. A ``summary edge'' in CFA2 or equivalently an
$\epsilon$-edge in PDCFA is an edge from the source of a push edge to
the target of the matching pop edge. ``Matching'' here means there is
a path through machine reductions that don't change the stack, or
through summary edges. In our model, we only ``push'' when we call a
function, and we only ``pop'' when we return with a value. There is an
analogy to something more operational: summary edges embody
memoization (see the last rule in figure
\ref{fig:summary-semantics}). Instead of following an entire path
through a call to return a value, we simply jump from the call to the
return with the result of the call (the second case of rule 4). We
will return to this analogy and what it means in the presence of
first-class control in section \ref{sec:sr}.

\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
% variable lookup
    $\tpl{(\svar\mvar, \menv), \mstore, \mkont, \mktab, \mmemo}$
    &
    $\tpl{\mval,\mstore,\mkont, \mktab, \mmemo}$ if $\mval \in \mstore(\menv(\mvar))$
    \\
% application
    $\tpl{(\sapp{\mexpri0}{\mexpri1}, \menv), \mstore, \mkont, \mktab, \mmemo}$
    &
    $\tpl{(\mexpri0, \menv), \mstore, \kar{\mexpri1,\menv,\mkont}, \mktab, \mmemo}$
    \\
% argument evaluation
    $\tpl{\mval, \mstore, \kar{\mexpr,\menv,\mkont}, \mktab, \mmemo}$
    &
    $\tpl{(\mexpr, \menv), \mstore, \kfn{\mval, \mkont}, \mktab, \mmemo}$
    \\
% function call
    $\tpl{\mval,\mstore,\kfn{\vclo{\slam{\mvar}{\mexpr}}{\menv}, \mkont},\mktab,\mmemo}$
    & % actually do call,
    $\tpl{\mpoint,
          \mstore',
          \krt{\mpoint, \mstore'},
          \joinone{\mktab}{(\mpoint, \mstore')}{\mkont},
          \mmemo}$
\\
    & or \\
    & $\tpl{\mval_\mathit{result}, \mstore', \mkont, \mktab, \mmemo} \text{ if } \mval_\mathit{result} \in \mmemo(\mpoint,\mstore')$
    \\ % or, lookup in table
%    & $\tpl{\mval, \mstore', \mkont,\mktab,\mmemo}$ if $\mval \in \mmemo(\mpoint, \mstore')$ \\    
    where & $\mpoint = (\mexpr, \extm{\menv}{\mvar}{\maddr})$ \\
          & $\mstore' = \joinone{\mstore}{\maddr}{\mval}$
    \\
% return
    $\tpl{\mval, \mstore, \krt{\mpoint,\mstore'}, \mktab, \mmemo}$
    &
    $\tpl{\mval, \mstore, \mkont, \mktab, \joinone{\mmemo}{(\mpoint, \mstore')}{\mval}}$
    if $\mkont \in \mktab(\mpoint, \mstore')$
  \end{tabular}
  \caption{The summarizing tabular stack machine}
  \label{fig:summary-semantics}
\end{figure}

@@

The immediately related work is that of PDCFA
\citep{ianjohnson:earl2010pdcfa},
CFA2~\citep{ianjohnson:vardoulakis-lmcs11,
  ianjohnson:Vardoulakis2011Pushdown}, and AAM
~\citep{dvanhorn:VanHorn2010Abstracting}, the first two of which we
recreated in full detail. The version of CFA2 that handles
\rackett{call/cc} does not handle composable control, and is dependent
on a restricted CPS representation. They also had no way to tune the
precision of their first class continuation approximation. Since
\rackett{call/cc} can be simulated with shift/reset, this work
supercedes theirs. The extended version of PDCFA that does garbage
collection \citep{dvanhorn:Earl2012Introspective} also fits into our
model, although we did not explicitly show it. The addresses that the
stack keeps alive can be accumulated by ``reading through'' the
continuation table, building up the set of addresses in each portion
of the stack that we come across.

@@

A new temporal logic that itself understands well-bracketing of call
and returns for stating and validating propositions, NWTL
~\citep{ianjohnson:alur:nwtl:2007}, applies to visibly pushdown
systems, which can precisely model progamming languages that only have
well-bracketed call and returns. The propositional form of the logic
has decidable satisfiability checking, and control-flow queries are
shallow propositions to pose in the logic. However, the satisfiability
problem is \textsc{2Exptime}-hard and the algorithm is not given in
the on-the-fly form that a higher-order language would need to
effectively answer control-flow queries without already knowing the
answers. The logic itself is an exciting new frontier for expressing
program correctness properties, and a higher-order version would be a
welcome addition to the functional programmer's toolbelt.

The trend of deriving static analyses from abstract machines does not
stop at flow analyses. The model-checking community showed how to
check temporal logic queries for collapsible pushdown automata (CPDA),
or equivalently, higher-order recursion schemes, by deriving the
checking algorithm from the Krivine
machine~\citep{ianjohnson:Salvati:2011:KMH:2027223.2027239}. The
expressiveness of CPDAs outweighs that of PDAs, but it is unclear how
to adapt higher-order recursion schemes to model arbitrary programming
language features. The method is strongly tied to the simply-typed
call-by-name lambda calculus and depends on finite sized base-types.

@@

Our goal in the future is to show that this technique is even more
widely applicable than shift and reset. We conjecture that the same
recipe will apply to the most intricate control operators in
production languages such as in
\citet{ianjohnson:Flatt:2007:ADC:1291151.1291178}. The control
structures there are difficult to model even with AAM's techniques due
to the ability to capture and compose continuations with arbitrarily many prompts,
but once we can tackle AAM, there should be a straightforward route to
extending it with the pushdown techniques of this paper.

There is also the question of clients of these analyses. Most
obviously we would want to know where we can implement first-class
control more efficiently. In particular, we would want to have an
escape analysis to find first-class continuations that don't need to
be heap-allocated, and single target abort operations that can be
turned into long jumps after a computed stack unraveling. On top of
optimizations, there are security analyses. Greater control of the
stack gives us the ability to drill deep into context-sensitive
security properties and make precise predictions such as, ``is it
possible to send image data on the network outside the context of the
user's permission?'' This way a user can use an image taking/sharing
application without the application surrepticiously taking and sending
images the user doesn't know about.
